Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-18 draft.tex
.

As an additional test, we applied external poisson input for $\mathrm{t\geq 300\, s}$ to two excitatory subgroups to see whether changes in the firing rate of these subgroups is levelled out by homeostasis or retained. Members of these subgroups were picked randomly and independent of their position.
\begin{figure}
\begin{center}
%% plot_external_input_switch_activity.py
\includegraphics[width=0.7\textwidth]{../../plots/external_input_switch.png}
\end{center}
\caption{Excitatory population firing rate of subgroups subject to external poisson input for $\mathrm{t\geq 300\, s}$. Red and green subgroups consist of 100 randomly picked exc. neurons each. The remaining 300 neurons (blue) received no external input.}
\label{Ext_Input_Switch}
\end{figure}

@@

A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, simulation-wise this case is quite easy to simulate: one simply has  to feed all NO-sources into a single scalar variable of NO concentration. In particular, this will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time, only shifting their initial random distribution. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory was $\mathrm{\sigma_{inst} = 1.41\, Hz}$ and the skewness $\mathrm{v_{inst} = 1.27}$ ($\mathrm{v_{inst,log} = -0.46}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 2 simulations and $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

@@

\subsection{The Firing Rate Distribution}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics, we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches to understand the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behaviour of recurrent networks are largely based on the assumption that the excitability of neurons is not individually tuned but rather adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this "canonical" approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal much differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behaviour.

Sweeney et al. claim that their implementation of diffusive homeostasis enables individual neurons to vary their firing rates more flexible compared to single-neuron homeostatic models, especially allowing for sustained encoding of different levels of external input. Interestingly, we could not confirm this, at least within the tests that we performed and the given set of standard parameters. Overall, the resulting behaviour of the excitatory population rather appeared to us as a form of single-neuron homeostasis whose individual targets are predetermined by the quenched random spatial structure of the respective cells. Of course, this impression only applies to the case where the aforementioned mixture of influence between spatial and synaptic topology is dominated by the former. In particular, our result depicted in Figure \ref{Ext_Input_Switch} did not support the notion of separability of inputs. Furthermore, by analyzing network activity and comparing it to our analytic model, we found that only one particular homeostatic fixed point of network activity is exists.

For the standard choice of parameters (being roughly based on experimental data), the resulting activity is indeed strongly bound to the spatial structure of the cells. Naturally, this raises the question of biological plausibility. To our knowledge, no experimental study exists that relates measurements of spontaneous activity to local fluctuations of neuronal density. Ivenshitz and Segal studied the influence of (mean) neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, since our network was tuned to exhibit non-bursting activity, it is hard to draw any conclusions from this result.

@@

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Since this effect is directly linked to the low density-high firing rate-relation shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (C), all previous thoughts regarding plausibility similarly apply to this result: \emph{If} a relationship between density and firing rates existed, we could indeed argue to also expect a relation as shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (D).

  If diffusion in a real biological system happens at a finite but fast speed compared to the adaptation of intrinsic excitability, one could arguably expect it to be equivalent to instant diffusion. In our simulation, we explained the fact that we saw a steady decline of correlation in Figure \ref{Corr_Coeff_vs_D} by the decreasing capability of the homeostatic mechanism to quickly enough react to structural synaptic changes that alter the recurrent excitatory inputs and, in consequence, excitatory firing rates.

A major problem of the implemented model is the usage of quasi-point sources of nitric oxide, since it leads to a clear single homeostatic fixed point. Future modifications of the homeostatic model might integrate a wider-spread kernel of nNOS, allowing for a larger amount of states of activity fulfilling the homeostatic target. In fact, experimental and theoretical studies have suggested that NO is synthesized across wider spread cell structures, leading to a more even distribution of NO \cite{Philippides_2005}.

While the implemented mathematical description of diffusion (Equation \eqref{NO_dyn}) is physically based, one can question the validity of the implementation of threshold adaption (Equation \eqref{Theta_dyn}): In principle, it does not impose any bounds onto the threshold value, which, from a biological point of view, seems implausible.

>>>>

As an additional test, we applied external poisson input for $\mathrm{t\geq 300\, s}$ to two excitatory subgroups to see whether changes in the firing rate of these subgroups is levelled out by homeostasis or retained. Members of these subgroups were picked randomly and independent of their position. As shown in Figure \ref{Ext_Input_Switch}, differences in among subgroups caused by different input signal strengths are gradually eliminated by the homeostatic mechanism.
\begin{figure}
\begin{center}
%% plot_external_input_switch_activity.py
\includegraphics[width=0.7\textwidth]{../../plots/external_input_switch.png}
\end{center}
\caption{Excitatory population firing rate of subgroups subject to external poisson input for $\mathrm{t\geq 300\, s}$. Red and green subgroups consist of 100 randomly picked exc. neurons each. The remaining 300 neurons (blue) received no external input.}
\label{Ext_Input_Switch}
\end{figure}

@@

A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, simulation-wise this case is quite easy to simulate: one simply has  to feed all NO-sources into a single scalar variable of NO concentration. In particular, this will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time, only shifting their initial random distribution. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory population was $\mathrm{\sigma_{inst} = 1.79\, Hz}$ and the skewness $\mathrm{v_{inst} = 1.51}$ ($\mathrm{v_{inst,log} = -0.52}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 10 simulations and $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

@@

An interesting idea that the aforementioned paper discusses is the possibility that slowness of homeostatic regulation on the timescale of hours might not only be due to functional reasons, such as this rate of adaption being sufficient to compensate for naturally occurring external perturbations. Based on their analysis, it might also be a mere issue of stability. In the context of this notion, faster homeostatic controls would simply be ruled out because they do not allow the nervous system to stay in a healthy condition. Our findings regarding the stability of diffusive homeostasis thus serves as an example supporting this hypothesis.

\subsection{The Firing Rate Distribution}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics, we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches of understanding the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behaviour of recurrent networks are largely based on the assumption that the excitability of neurons is not individually fine-tuned but rather randomly distributed and adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this "canonical" approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal many differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behaviour.

Sweeney et al. claim that their implementation of diffusive homeostasis enables individual neurons to vary their firing rates more flexible compared to single-neuron homeostatic models, especially allowing for sustained encoding of different levels of external input. Interestingly, we could not confirm this, at least within the tests that we performed and the given set of standard parameters. Overall, the resulting behaviour of the excitatory population rather appeared to us as a form of single-neuron homeostasis whose individual targets are predetermined by the quenched random spatial structure of the respective cells. In particular, our result depicted in Figure \ref{Ext_Input_Switch} did not support the notion of separability of inputs. Of course, this impression only applies to the case where the aforementioned mixture of influence between spatial and synaptic topology is dominated by the former. By analyzing network activity based on our analytic model, we found that for finite diffusion constants in theory, indeed only a single homeostatic fixed point exists. On a practical level though, the gradual decorrelation with this theoretical state of activity upon larger values of $\mathrm{D}$ is presumably caused by synaptic turnovers, which change the synaptic input. The resulting changes in activity are more slowly and less accurately compensated in the case of larger diffusion constants.

However, for the standard choice of parameters (being roughly based on experimental data), the resulting activity is indeed strongly bound to the spatial structure of excitatory cells. Naturally, this raises the question of biological plausibility. To our knowledge, no experimental study exists that relates measurements of spontaneous activity to local fluctuations of neuronal densities. Ivenshitz and Segal studied the influence of (mean) neuronal density onto neuronal activity in a culture of hippocampal neurons \cite{Ivenshitz_2010}. Lower densities resulted in a more irregular activity, with stronger but less frequent bursting events. However, since our network was tuned to exhibit non-bursting activity, it is hard to draw any conclusions from this result.

@@

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Since this effect is directly linked to the low-density/high-firing rate-relation shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (C), all previous thoughts regarding plausibility similarly apply to this result: \emph{If} a relationship between density and firing rates existed, we could indeed argue to also expect a relation as shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (D), since it is known that neurons above-average firing rates tend to undergo long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}.

\subsubsection{Possible Modifications}
Since these considerations evoked a rather sceptical view onto the biological plausibility of these results, one should consider possible modifications of the homeostatic model. One should note that these are to be evaluated in the context of providing a model of an actual physiological process. If the main goal was to simply provide a homeostatic method that retains an overall level of activity without tightly fixing individual rates, we would suggest to use the idealized model of instant diffusion, especially because of its computational simplicity. However, as already mentioned, the given experimentally based parameters do not allow to accept this simplification from the standpoint of biological and physical accuracy. Two elements of the model should be put into question: First, is it appropriate to assume an approximately point-like source of nitric oxide? Second, are the dynamics of threshold adaption (Equation \eqref{Theta_dyn}) an acceptable description of the actual physiological processes?

While Sweeney et al. argue that the size of the studied tissue is large compared to the size of individual somata (which, apparently, they claim to be the main source of nitric oxide), Philippides et al. claim that NO is often synthesized in a more delocalized way by means of fine fibres exceeding the soma \cite{Philippides_2005}. The diameter of these "production areas" is on the order of $\mathrm{10-100 \mu m}$. To incorporate this spread of synthesis in our model of diffusion, one would - for each neuron - need to convolve a suitably shaped and shifted "production"-kernel (scaled to the momentary rate of NO-synthesis) with the field of NO-concentration. Intuitively, this would locally "smear" the distinction between neighbouring neurons. From a more abstract point of view, this would provide a way to drive the matrix defined in Equation \eqref{NO_equil_cond_interact_matrix_operator} towards singularity, allowing for more states of activity to approximately fulfil the homeostatic target.

Judging the plausibility of the used model of threshold adaptation is a rather problematic task. Experimental studies have investigated the effect of NO onto the conductance of ionic channels \cite{Steinert_NO,Steinert_NO_2011}. Mapping these features onto a simple LIF-model is obviously not easy, and many consequences of NO-modulation onto neural dynamics can certainly not be represented in such a simple model. Our criticism is thus rather based on more general arguments.

A major issue of Equation \eqref{Theta_dyn} is the fact that, given a clamped positive or negative difference between the actual NO concentration and the target concentration, the excitability of the neuron would decrease or increase without bounds. Because of the limited amount of ion channels that exist in a cell and can be affected by nitric oxide, upper and lower bounds should certainly exist. A simple approach to tackle this would be to implement hard bounds into our neuron model, not allowing the threshold to pass an upper and lower value. Another possibility could be to implement these bounds by means of the differential equation itself.

In either case, such a modification would break the ability of the network to tune thresholds in such a way that \emph{all} neurons eventually reach target concentration.
